DOpElib: A Goal Oriented Software Library for Computing PDEs and Optimization Problems

### 2012-10-15 Paper_May_2012/main.tex
Paper: Nonstationary Problem Beschreibung angepasst und auf Wunsch von Christian die Funktionalauswertung in dem FSI Beispiel hinzugefuegt, so dass klar ist, was die vier Graphen aussagen

\subsubsection{Nonstationary problems}
In this section, we focus on the extension to 
the numerical solution of nonstationary processes. 
As example, we consider the heat equation:
Find $u$:
\[
\partial_t u - \Delta u = 0
\]
and in weak formulation:
Find $u$:
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$ and 
\[
a(u)(\phi) = (\partial_t u, \phi) + (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
Temporal discretization using the One-step-$\theta$ scheme reads:
Given the old timestep solution $u^n$, find $u^{n+1}$:
\[
\frac{(u^{n+1} - u^{n}, \phi)}{k} + \theta (\nabla u^{n+1}, \nabla \phi)
+ (1 - \theta) (\nabla u^{n}, \nabla \phi) = 0,
\]
Reordering in new and old terms and multiplication by the time step size $k$:
\[
(u^{n+1},\phi) + k \theta (\nabla u^{n+1}, \nabla \phi)
= (u^n, \phi) - (1 - \theta) (\nabla u^{n}, \nabla \phi).
\]
In \dope, different time stepping schemes can be found 
in the folder \texttt{DOpEsrc/tsschemes}.

Because the \texttt{main} file 
is basically the same as for stationary problems
which is one major goal of the software library,
we only focus on the implementation of the PDE in the 
\texttt{localpde.h} file. Here, the stationary part, i.e.,
the diffusion is implemented as before. The only thing we add 
is the time-depdendent part $\partial_t u$:

\begin{lstlisting}
  void CellTimeEquation(...)
   {
     // implement (u^{n+1},\phi) 
   }
\end{lstlisting}
and it is done in a straight-forward way:
\begin{lstlisting}
  for (all quadrature points per cell q)
   {
     Tensor<1,2> u = ...;

@@

        local_cell_vector(i) +=  
          scale * (u * phi_i_u) * 
          state_fe_values.JxW(q_point);
      }
   }
\end{lstlisting}
Here, the advantage is that \dope orders the terms in the timestepping
problems
such that only $(u^{n+1},\phi)$ must be implemented. However, if someone 
needs explicit input of the old timestep term, then he can use 
\begin{lstlisting}
  void CellTimeEquationExplicit (...)
   {
     // implement (u^{n+1},\phi) and (u^n,\phi)  
   }
\end{lstlisting}
For instance, this explicit term is used for fluid-structure interaction in
which
the old timestep term needs explicit modification.

@@

@@

\begin{figure}
\centering
{\includegraphics[width=5.8cm]{Pictures/ux_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=5.8cm]{Pictures/uy_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=5.8cm]{Pictures/Drag_fluid_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=5.8cm]{Pictures/Lift_fluid_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
\caption{FSI 2: the deflections of the beam, $u_x(A)$ and $u_y(A)$ (in $cm$), and 
the drag and the lift evaluation (in $kg/m\,s^2$) are displayed versus
time (in $s$).
} 
\label{res:results_ux_and_uy_fsi_2}
\end{figure}

>>>>

\subsubsection{Nonstationary problems}
In this section, the 
numerical solution of nonstationary processes
is considered. We focus on details which are 
beyond stationary examples as discussed in the 
previous sections. Before, we introduce 
the steps of implementation in DOpElib, we 
derive the basic theoretical concepts.
As an example, we consider the heat equation:
Find $u\in V$:
\[
\partial_t u - \Delta u = 0
\]
and in weak formulation:
Find $u\in V$:
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$ and 
\[
a(u)(\phi) = (\partial_t u, \phi) + (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
Temporal discretization using the One-step-$\theta$ scheme reads:
Given the old timestep solution $u^n$, find $u^{n+1}$:
\[
\frac{(u^{n+1} - u^{n}, \phi)}{k} + \theta (\nabla u^{n+1}, \nabla \phi)
+ (1 - \theta) (\nabla u^{n}, \nabla \phi) = 0,
\]
Reordering in new and old terms and multiplication by the time step size $k$:
\[
(u^{n+1},\phi) + k \theta (\nabla u^{n+1}, \nabla \phi)
= (u^n, \phi) - (1 - \theta) (\nabla u^{n}, \nabla \phi).
\]
The concrete choice of $\theta$ leads to a
specific time stepping scheme. For instance,
$\theta = 0.5$ gives the Crank-Nicolson scheme,
\begin{lstlisting}
   CrankNicolsonProblem<...>
\end{lstlisting}
The main work is treated as for stationary problems
within 
\begin{lstlisting}
   MethodOfLines_SpaceTimeHandler<...>
\end{lstlisting}
Again, the key issue is that this procedure is totally independent 
of the specific form of the PDE. The user-related work 
remains to provide the time derivative $\partial_t u$ in
the implementation:
\begin{lstlisting}
  void CellTimeEquation(...)
   {
     // implement (u^{n+1},\phi) 
   }
\end{lstlisting}
and it is done in a straight-forward way:
\begin{lstlisting}
  for (all quadrature points per cell q)
   {
     Tensor<1,2> u = ...;

@@

        local_cell_vector(i) +=  
          scale * (u * phi_i_u) * 
          state_fe_values.JxW(q_point);
      }
   }
\end{lstlisting}
Here, the advantage is that \dope orders the terms in the timestepping
problems
such that only $(u^{n+1},\phi)$ must be implemented. However, if someone 
needs explicit input of the old timestep term, then he can use 
\begin{lstlisting}
  void CellTimeEquationExplicit (...)
   {
     // implement (u^{n+1},\phi) and (u^n,\phi)  
   }
\end{lstlisting}

@@

The quantities of interest are evaluations of 
$x$- and $y$ displacement at the point $A(0) = (0.6,0.2)$
and the drag and lift forces acting on the cylinder and the elastic beam:
\begin{align}
\label{drag_lift_forces}
(F_D , F_L) 
= {\int_{S_f} \sigma_f \cdot n_f \, ds + 
\int_{\Gamma_i} \sigma_s \cdot n_s \, ds},
\end{align}
where $S_{f}$ denotes the path over the cylinder in the fluid part and
$\Gamma_i$ the interface between the elastic beam and the 
fluid.

@@

\begin{figure}
\centering
{\includegraphics[width=5.8cm]{Pictures/ux_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=5.8cm]{Pictures/uy_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=5.8cm]{Pictures/Drag_fluid_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
{\includegraphics[width=5.8cm]{Pictures/Lift_fluid_FSI_2_FS_t_3e-2_t_15e-3_global_2_Hron_grid.pdf}}
\caption{FSI 2: the deflections of the beam, $u_x(A)$ and $u_y(A)$ (in $cm$), and 
the drag $F_D$ and the lift $F_L$ evaluation (in $kg/m\,s^2$) are displayed versus
time (in $s$).
} 
\label{res:results_ux_and_uy_fsi_2}
\end{figure}
